Optimal updating factor in adaptive flat-distribution simulations

### 2016-04-22 doc/invt.tex
invtrun.py -> invtcscan.py, add log(c) scan with --x
improve the code for predicting error components

As shown above, with a fixed value of $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
%\notag
\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
\notag
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
which coincides with the Lagrangian,
is conserved.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  \left(
    \frac C T
  \right)^2
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C$.
%
Integrating the square root, we get
%
\begin{equation}
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  C \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  C =
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
Since we only optimized the asymptotic error, $E_A$,
in th above,
the parameter $q(T)$ can be further adjusted
to minimize the total error, $E = E_R + E_A$.
%
A complete program of computing the optimal schedule
is listed in Appendix \ref{sec:program}.

@@

We can characterize the optimal schedule
geometrically as follows.
%
First, we define the mass function as
$$
  M(Q)
  =
  \sqrt{
    \textstyle\sum_{ k = 2 }^n \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, Q}
  }
  .
$$
and the normalized mass distribution as
\begin{equation}
  m(Q)
  =
  \frac{
    M(Q)
  }
  {
    \int_0^{ q(T) } M(Q') \, d Q'
  }
  =
  \frac{
    M(Q)
  }
  {
    C
  }
  ,
\label{eq:mass_distr}
\end{equation}
%
such that
$\int_0^{q(T)} m(Q) \, dQ = 1$.
%
Clearly,
the distribution decreases monotonically with $Q$.

@@

Then, from Eq. \eqref{eq:q_opt},
we find that for $Q = q(T) - q(t)$,
%
\begin{align}
  m(Q)
  &=
  \frac{ 1 }
       { T \, \alpha(t) }
  ,
\label{eq:mQ_invTa}
  \\
  \int_Q^{ q(T) }
    m(Q') \, dQ'
  &=
  \frac t T
  ,
\label{eq:intmQ_tT}
\end{align}
%
which means that each point $(t, \alpha(t))$
on the schedule can be mapped
to a point $(Q, m(Q))$ on the mass distribution,
such that the ordinate of the latter curve
gives $1/(T\alpha)$,
and the area under the curve in the domain $[Q, q(T)]$
is equal to $t/T$,
as shown in Fig. \ref{fig:massq},

@@

If $\mu_2 = \dots = \mu_b = 0$,
the matrix $\mathbf w$ is tridiagonal,
%%
%\begin{equation}
%\arraycolsep=3.6pt\def\arraystretch{1.4}
%\mathbf w
%=
%\left(
%  \begin{array}{cccccccc}
%    1 - \mu_1   & \mu_1 & 0 & \dots & 0 \\
%    \mu_1 & 1 - 2 \, \mu_1  & \mu_1 & \dots & 0 \\
%    \vdots & &  & & \vdots \\
%    0 & \dots & \mu_1 & 1 - 2 \, \mu_1  & \mu_1 \\
%    0 & \dots & 0 & \mu_1 & 1 - \mu_1
%  \end{array}
%\right).
%\label{eq:wnn}
%\end{equation}
%%
%We have
and the eigenvalues are
\begin{equation}
  \lambda_k
  =
  1 -
  4 \, \mu_1 \sin^2
  \frac{ (k - 1) \, g \, \pi }
       {       2 \, n        }
  .
\notag
%\label{eq:wnn_eigenvalue}
\end{equation}
%
Thus, this updating scheme is stable if
$\min \lambda_k = \lambda_n \ge 0$,
or
\begin{equation}
  \mu_1 \le
  \begin{dcases}
    \frac 1 4
    & \mathrm{periodic \; and \;} n \mathrm{ \; even,}
    \\
    \frac 1 4
    \sec^2
    \left( \frac{  \pi   }
                { 2 \, n }
    \right)
    & \mathrm{otherwise}.
  \end{dcases}
\notag
%\label{eq:nn_stable}
\end{equation}

@@

For $n \gg 1$,
we define
$x = l \, g \, \Delta x$,
and
$\mu_l = \mu(x) \, \Delta x$,
where
$\Delta x = \pi/n$
is the bin size.
%
Then,
by assuming the largest $b \approx n/g$,
Eq. \eqref{eq:wband_eigenvalue}
can be approximated as an integral:
%
\begin{equation}
  \lambda_k
  =
  2 \int_0^\pi
    \mu(x) \, \cos \bigl[ (k-1) \, x \bigr] \, dx,
\notag
%\label{eq:lambda_int}
\end{equation}
%
with the normalization
%
$
  1 = 2 \int_0^\pi \mu(x) \, dx.
$

For the Gaussian updating kernel of width $\sigma$
with $\Delta x \ll \sigma \ll 1$,
we can extended
the upper limit of the integrals
to infinity, and
%
\begin{equation}
  \mu(x)
  =
  \frac{            1            }
       { \sqrt{ 2 \pi \sigma^2 } }
  %
  \exp\left(
        -\frac{       x^2     }
              { 2 \, \sigma^2 }
      \right)
  .
\notag
\end{equation}
%
%
If follows that all eigenvalues are positive:
%
\begin{align}
  \lambda_k
  &=
  \exp\left[
        -\frac{ (k - 1)^2 \, \sigma^2 }
              {           2           }
      \right]
  %\notag
  %\\
  %&=
  %\exp\left[
  %      -\frac{ \pi^2 }{ 2 }
  %      \left(
  %        \frac{ k - 1 }
  %             {   n   }
  %      \right)^2
  %      \left(
  %        \frac{  \sigma }
  %             { \Delta x }
  %      \right)^2
  %    \right]
  .
\notag
%\label{eq:lambda_Gaussian}
\end{align}

@@

We shall first show
that the single-bin scheme is asymptotically optimal.
%
Consider a class of $\mathbf w$ matrices
sharing the same set of eigenvectors,
hence the same $\Gamma_k$'s,
but with different positive eigenvalues,
$\lambda_k$'s.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of nonnegative numbers, $c_k$'s,
%
\begin{align}
&
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr)
\right)
%
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, c_k^2
\right)
%
\notag
\\
&
\qquad \qquad
\ge
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, c_k \, \dot u_k \, \bigl( q(t) \bigr)
\right)^2
\notag
\\
&
\qquad \qquad
=
\left(
  \sum_{k = 2}^n \Gamma_k \, c_k
    \left[
      1 - e^{ -\lambda_k \, q(T) }
    \right]
\right)^2.
\label{eq:CSineq}
\end{align}
%
\note{This inequality follows from the non-negativity of
the following quadratic polynomial of $Y$,
$$
\int_0^T
  dt \sum_{k = 2}^n \Gamma_k \,
    \left( \dot u_k \, Y - c_k \right)^2
  \equiv
  A \, Y^2 + B \, Y + C
  ,
$$
which implies a non-positive discriminant,
$B^2 - 4 \, A \, C$.
}%
%
Note that the last expression of \eqref{eq:CSineq}
depends on the curve, $q(t)$ ($0 < t < T$),
only through the endpoint value, $q(T)$,
which is fixed in the variational process.
%
Thus, the inequality sets a lower bound
for the asymptotic error $\Err$
given in Eq. \eqref{eq:error_asym1}.
%
The equality is achieved
if $\dot u_k\left( q(t) \right) = c_k$
for all $k = 2, \dots, n$ at any $t$,
up to a multiplicative factor.
%
Solving these equations yields
%
\begin{equation}
  \alpha(t) = \frac{              1             }
                   { \lambda_k \, (t + t_{k0} ) },
\label{eq:alpha_invtk}
\end{equation}
with
\begin{equation}
  t_{k0} = \frac{             T            }
                { e^{ \lambda_k q(T) } - 1 }.
\label{eq:tk0}
\end{equation}
%
Such a solution is possible only if
%
\begin{equation}
  \lambda_2 = \dots = \lambda_n.
  \notag
\end{equation}
%
Then, by Eq. \eqref{eq:tk0},
all $t_{k0}$'s also share the same value,
$t_0$,
such that
$u_k = (t + t_0) / (T + t_0)$,
and
$c_k = 1/(T + t_0)$
for $k \ge 2$.
%
The asymptotic error is then
\begin{align}
  E_A
  =
  \frac{       T     }
       { (T + t_0)^2 }
  \sum_{ k = 2 }^n
    \Gamma_k
  .
\label{eq:error_singlebin}
\end{align}
%
If we further assume $\lambda_2 = 1$,
%the optimal schedule given by
Eq. \eqref{eq:alpha_invtk}
recovers Eq. \eqref{eq:alpha_invt1}.
%
\note{We add some steps for the above solution.
  Integrating $\dot u_k = c_k$ yields
  \begin{equation}
    u_k(t) = c_k \left(t + t_{k0} \right).
  \tag{N4}
  \label{neq:ui_solution}
  \end{equation}
  Given that $u_k(T) = 1$ and $u_k(0) = e^{-\lambda_k \, q(T)}$,
  we have
  $$
  c_k = \frac{ 1 }{ T + t_{k0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{k0} } { T + t_{k0} }
  =
  e^{ -\lambda_k \, q(T) }.
  $$
  Taking the logarithm of Eq. \eqref{neq:ui_solution} yields
  $$
  -\lambda_k \, q(T) + \lambda_k \, q(t)
  = \ln c_k + \ln \left( t + t_{k0} \right).
  $$
  Differentiating this with respect to $t$ yields
  Eq. \eqref{eq:alpha_invtk}.
}

@@

We can generalize
the single-bin scheme to a class of
perfect (brick-wall) \emph{bandpass} updating schemes
to allow zero eigenvalues.
%
For example,
if we can assume that the PMF is sufficiently smooth
such that the eigenmodes can be limited to $k \le K$.
%
Then we can construct an updating scheme with
$$
\lambda_1 = \cdots = \lambda_K = 1,
\mathrm{\; and \;}
\lambda_{K+1} = \cdots = \lambda_n = 0.
$$
%
Then,
Eq. \eqref{eq:error_singlebin} is changed to
%
\begin{equation}
  E_A
  =
  \frac {       T     }
        { (T + t_0)^2 }
  \sum_{ k = 2 }^K
    \Gamma_k.
\label{eq:error_asym_sinc}
\end{equation}
%
The optimal schedule of these schemes
remains to be the inverse-time one
given by Eq. \eqref{eq:alpha_invt1}.
%
Below, we construct these schemes for
the homogeneous case.

@@

In our test system,
$i$ is the index of a one-dimensional
non-periodic variable,
and $n = 100$.
%
Both the target and intrinsic distributions
are assumed to be flat:
$p^*_i = p_i = 1/n$.
%
%We wish to see how
%the schedule, $\alpha(t)$,
%affects the fluctuating error
%of the bias potential
%at the end of the simulation.
%
Initially,
we set the bias potential to zero,
$V_i \equiv 0$,
%
and equilibrate the system for $10^7$ steps
under a constant updating magnitude,
$\alpha_0 = 10^{-4}$,
unless specified otherwise.
%
After this,
we reset the origin of the time, $t$,
and start the testing schedule.
%specified by either
%Eq. \eqref{eq:q_opt}
%or
%Eq. \eqref{eq:alpha_invtlambda}.

@@

We tested the theory of computing errors
on the simple inverse-time schedule,
%
\begin{equation}
\alpha(t) = \frac{1}{\lambda \, (t + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
where $\lambda$ is a free parameter,
and $t_0 = 2/\alpha_0$.
%
The error of this schedule
can be computed analytically
(cf. Appendix \ref{sec:invt_schedule}),
and tested against simulation.

We performed FDS simulations
using the single-bin updating scheme
and the triple-bin updating scheme with $\mu_1 = 0.24$.
%
For each value of $\lambda$, we ran $1000$ independent simulations
of $10^8$ steps, and averaged the results.
%
As shown in Fig. \ref{fig:errinvt},
the errors from the simulations
agreed well those from the theory.

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsigscan.pdf}
  }
  \caption{
    \label{fig:errsigscan}
    Normalized error, $(T + t_0) \, E$,
    versus the width, $\sigma$,
    for the Gaussian
    and the bandpass updating schemes.
    %
    In the latter case,
    $\sigma$ is given by Eq. \eqref{eq:sigma_equiv}.
    %
    Here, we assumed $t_0 \equiv T/(e^{q(T)} - 1)$
    for the optimal value of $q(T)$.
    %
    The points were obtained from averaging results
    from 1000 independent simulations;
    the curves are theoretical predictions.
    %
    \note{This figure was produced by
      \texttt{doc/fig/errsigscan.gp}.
      The data were saved under
      \texttt{data/scan}.
    }%
  }
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/gamcmp.pdf}
  }
  \caption{
    \label{fig:gamcmp}
    Integrals of the autocorrelation functions
    from an MD simulation versus those
    from the one-step process.
    %
    The MD simulation was ran for $10^{10}$ steps with
    the step size $\Delta t_\mathrm{MD} = 0.005$.
    %
    The bin size $\Delta z = 2\, \pi /n$ with $n = 100$,
    and the temperature was $1.0$.
    %
    The velocity rescaling\cite{bussi2007} thermostat was
    applied every MD step with viscocity 20.
    %
    The reflective boundary condition was used.
    %
    \note{The figure was produced by \texttt{doc/fig/gamcmp.gp}.
      Thee data were saved under \texttt{data/gamcmp}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}

@@

For the single-bin scheme, we have
$\lambda_2 = \cdots = \lambda_n = 1$,
and
$\Gamma_k$'s affect the errors
%from Eqs. \eqref{eq:error_res_invt}
%and \eqref{eq:error_asym_invt}
only via the sum, $\sum_{k = 2}^n \Gamma_k$,
as a multiplicative constant.

>>>>

As shown above, with a fixed value of $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
%\notag
\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
\notag
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
which coincides with the Lagrangian,
is conserved.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  \left(
    \frac C T
  \right)^2
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C$.
%
Integrating the square root, we get
%
\begin{equation}
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  C \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  C =
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
Since we only optimized the asymptotic error, $E_A$,
in the above,
the parameter $q(T)$ can be further adjusted
to minimize the total error, $E = E_R + E_A$.
%
A complete program of computing the optimal schedule
is listed in Appendix \ref{sec:program}.

@@

We can characterize the optimal schedule
geometrically as follows.
%
First, we define the mass function as
$$
  M(Q)
  =
  \sqrt{
    \textstyle\sum_{ k = 2 }^n \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, Q}
  }
  .
$$
and the normalized mass distribution as
\begin{equation}
  m(Q)
  =
  \frac{
    M(Q)
  }
  {
    \int_0^{ q(T) } M(Q') \, d Q'
  }
  =
  \frac{
    M(Q)
  }
  {
    C
  }
  ,
\notag
%\label{eq:mass_distr}
\end{equation}
%
such that
$\int_0^{q(T)} m(Q) \, dQ = 1$.
%
Clearly,
the distribution decreases monotonically with $Q$.

@@

Then, from Eq. \eqref{eq:q_opt},
we find that for $Q = q(T) - q(t)$,
%
\begin{align}
  m(Q)
  &=
  \frac{ 1 }
       { T \, \alpha(t) }
  ,
\label{eq:mQ_invTa}
  \\
  \int_Q^{ q(T) }
    m(Q') \, dQ'
  &=
  \frac t T
  ,
\label{eq:intmQ_tT}
\end{align}
%
which means that each point
$\bigl(t, \alpha(t)\bigr)$
on the schedule can be mapped
to a point
$\bigl(Q, m(Q)\bigr)$
on the mass distribution,
such that the ordinate of the latter curve
gives $1/(T\alpha)$,
and the area under the curve in the domain $[Q, q(T)]$
is equal to $t/T$,
as shown in Fig. \ref{fig:massq},

@@

If $\mu_2 = \dots = \mu_b = 0$,
the matrix $\mathbf w$ is tridiagonal,
%%
%\begin{equation}
%\arraycolsep=3.6pt\def\arraystretch{1.4}
%\mathbf w
%=
%\left(
%  \begin{array}{cccccccc}
%    1 - \mu_1   & \mu_1 & 0 & \dots & 0 \\
%    \mu_1 & 1 - 2 \, \mu_1  & \mu_1 & \dots & 0 \\
%    \vdots & &  & & \vdots \\
%    0 & \dots & \mu_1 & 1 - 2 \, \mu_1  & \mu_1 \\
%    0 & \dots & 0 & \mu_1 & 1 - \mu_1
%  \end{array}
%\right).
%\label{eq:wnn}
%\end{equation}
%%
%We have
and the eigenvalues are
\begin{equation}
  \lambda_k
  =
  1 -
  4 \, \mu_1 \sin^2
  \frac{ (k - 1) \, g \, \pi }
       {       2 \, n        }
  .
\notag
%\label{eq:wnn_eigenvalue}
\end{equation}
%
So this updating scheme is stable if
$\min \lambda_k = \lambda_n \ge 0$,
or
\begin{equation}
  \mu_1 \le
  \begin{dcases}
    \frac 1 4
    & \mathrm{periodic \; and \;} n \mathrm{ \; even,}
    \\
    \frac 1 4
    \sec^2
    \left( \frac{  \pi   }
                { 2 \, n }
    \right)
    & \mathrm{otherwise}.
  \end{dcases}
\notag
%\label{eq:nn_stable}
\end{equation}

@@

For $n \gg 1$,
we define
$x = l \, g \, \Delta x$,
and
$\mu_l = \mu(x) \, \Delta x$,
where
$\Delta x = \pi/n$
is the bin size.
%
By assuming the largest $b \approx n/g$,
Eq. \eqref{eq:wband_eigenvalue}
can be approximated as an integral:
%
\begin{equation}
  \lambda_k
  =
  2 \int_0^\pi
    \mu(x) \, \cos \bigl[ (k-1) \, x \bigr] \, dx,
\notag
%\label{eq:lambda_int}
\end{equation}
%
with the normalization
%
$
  1 = 2 \int_0^\pi \mu(x) \, dx.
$

For the Gaussian updating kernel of width $\sigma$
with $\Delta x \ll \sigma \ll 1$,
we can extend
the upper limit of the integrals
to infinity, and
%
\begin{equation}
  \mu(x)
  =
  \frac{            1            }
       { \sqrt{ 2 \pi \sigma^2 } }
  %
  \exp\left(
        -\frac{       x^2     }
              { 2 \, \sigma^2 }
      \right)
  .
\notag
\end{equation}
%
%
If follows that all eigenvalues are positive:
%
\begin{align}
  \lambda_k
  &=
  \exp\left[
        -\frac{ (k - 1)^2 \, \sigma^2 }
              {           2           }
      \right]
  %\notag
  %\\
  %&=
  %\exp\left[
  %      -\frac{ \pi^2 }{ 2 }
  %      \left(
  %        \frac{ k - 1 }
  %             {   n   }
  %      \right)^2
  %      \left(
  %        \frac{  \sigma }
  %             { \Delta x }
  %      \right)^2
  %    \right]
  .
\notag
%\label{eq:lambda_Gaussian}
\end{align}

@@

We shall first show
that the single-bin scheme is asymptotically optimal.
%
Consider a class of $\mathbf w$ matrices
sharing the same set of eigenvectors,
hence the same $\Gamma_k$'s,
but with different positive eigenvalues,
$\lambda_k$'s.
%
Using the Cauchy-Schwarz inequality, we have,
for any set of nonnegative numbers, $c_k$'s,
%
\begin{align}
&
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr)
\right)
%
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, c_k^2
\right)
%
\notag
\\
&
\qquad \qquad
\ge
\left(
  \int_0^T dt
    \sum_{k = 2}^n
      \Gamma_k \, c_k \, \dot u_k \, \bigl( q(t) \bigr)
\right)^2
\notag
\\
&
\qquad \qquad
=
\left(
  \sum_{k = 2}^n \Gamma_k \, c_k
    \left[
      1 - e^{ -\lambda_k \, q(T) }
    \right]
\right)^2.
\label{eq:CSineq}
\end{align}
%
\note{This inequality follows from the non-negativity of
the following quadratic polynomial of $Y$,
$$
\int_0^T
  dt \sum_{k = 2}^n \Gamma_k \,
    \left( \dot u_k \, Y - c_k \right)^2
  \equiv
  A \, Y^2 + B \, Y + C
  ,
$$
which implies a non-positive discriminant,
$B^2 - 4 \, A \, C$.
}%
%
Note that the last expression of \eqref{eq:CSineq}
depends on the curve, $q(t)$ ($0 < t < T$),
only through the endpoint value, $q(T)$,
which is fixed in the variational process.
%
Thus, the inequality sets a lower bound
for the asymptotic error $\Err$
given in Eq. \eqref{eq:error_asym1}.
%
The equality is achieved
if $\dot u_k\left( q(t) \right) = c_k$
for all $k = 2, \dots, n$ at any $t$,
up to a multiplicative factor.
%
Solving these equations yields
%
\begin{equation}
  \alpha(t) = \frac{              1             }
                   { \lambda_k \, (t + t_{k0} ) },
\label{eq:alpha_invtk}
\end{equation}
with
\begin{equation}
  t_{k0} = \frac{             T            }
                { e^{ \lambda_k q(T) } - 1 }.
\label{eq:tk0}
\end{equation}
%
Such a solution is possible only if
%
\begin{equation}
  \lambda_2 = \dots = \lambda_n.
  \notag
\end{equation}
%
Then, by Eq. \eqref{eq:tk0},
all $t_{k0}$'s also share the same value,
$t_0$,
such that
$u_k = (t + t_0) / (T + t_0)$,
and
$c_k = 1/(T + t_0)$
for $k \ge 2$.
%
The asymptotic error is then
\begin{align}
  E_A
  =
  \frac{       T     }
       { (T + t_0)^2 }
  \sum_{ k = 2 }^n
    \Gamma_k
  .
\label{eq:error_singlebin}
\end{align}
%
If we further assume $\lambda_2 = 1$,
%the optimal schedule given by
Eq. \eqref{eq:alpha_invtk}
recovers Eq. \eqref{eq:alpha_invt1}.
%
\note{Derivation.
  Integrating $\dot u_k = c_k$ yields
  \begin{equation}
    u_k(t) = c_k \left(t + t_{k0} \right).
  \tag{UK}
  \label{neq:ui_solution}
  \end{equation}
  Given that $u_k(T) = 1$ and $u_k(0) = e^{-\lambda_k \, q(T)}$,
  we have
  $$
  c_k = \frac{ 1 }{ T + t_{k0} },
  \quad
  \mathrm{and\;}
  \frac{ t_{k0} } { T + t_{k0} }
  =
  e^{ -\lambda_k \, q(T) }.
  $$
  Taking the logarithm of Eq. \eqref{neq:ui_solution} yields
  $$
  -\lambda_k \, q(T) + \lambda_k \, q(t)
  = \ln c_k + \ln \left( t + t_{k0} \right).
  $$
  Differentiating this with respect to $t$ yields
  Eq. \eqref{eq:alpha_invtk}.
}

@@

We can generalize
the single-bin scheme to a class of
perfect (brick-wall) \emph{bandpass} updating schemes
to allow zero eigenvalues.
%
For example,
if we can assume that the PMF is sufficiently smooth
such that the eigenmodes can be limited to $k \le K$.
%
Then we can construct an updating scheme with
$$
\lambda_1 = \cdots = \lambda_K = 1,
\mathrm{\; and \;}
\lambda_{K+1} = \cdots = \lambda_n = 0.
$$
%
Then,
we can change the upper limit of the sum in
Eq. \eqref{eq:error_singlebin} to $K$.
%
%Then,
%Eq. \eqref{eq:error_singlebin} is changed to
%%
%\begin{equation}
%  E_A
%  =
%  \frac {       T     }
%        { (T + t_0)^2 }
%  \sum_{ k = 2 }^K
%    \Gamma_k.
%%\notag
%\label{eq:error_asym_sinc}
%\end{equation}
%
The optimal schedule of these schemes
remains to be the inverse-time one
given by Eq. \eqref{eq:alpha_invt1}.
%
Below, we construct these schemes for
the homogeneous case.

@@

In our test system,
$i$ represents a one-dimensional
non-periodic variable,
and $n = 100$.
%
Both the target and intrinsic distributions
are assumed to be flat:
$p^*_i = p_i = 1/n$.
%
%We wish to see how
%the schedule, $\alpha(t)$,
%affects the fluctuating error
%of the bias potential
%at the end of the simulation.
%
Initially,
we set the bias potential to zero,
$V_i \equiv 0$,
%
and equilibrate the system for $10^7$ steps
under a constant updating magnitude,
$\alpha_0 = 10^{-4}$,
unless specified otherwise.
%
After this,
we reset the origin of the time, $t$,
and start the testing schedule.
%specified by either
%Eq. \eqref{eq:q_opt}
%or
%Eq. \eqref{eq:alpha_invtlambda}.

@@

We tested the theory of computing errors
on the simple inverse-time schedule,
%
\begin{equation}
\alpha(t) = \frac{1}{\lambda \, (t + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
where $\lambda$ is a free parameter,
and $t_0$ is given by Eq. \eqref{eq:t0_sinc}.
%and $t_0 = 2/\alpha_0$.
%
The error of this schedule
can be computed analytically
(cf. Appendix \ref{sec:invt_schedule}),
and tested against simulation.

We performed adaptive FDS simulations
using the single-bin updating scheme
and the triple-bin updating scheme with $\mu_1 = 0.24$.
%
For each value of $\lambda$, we ran $1000$ independent simulations
of $10^8$ steps, and averaged the results.
%
As shown in Fig. \ref{fig:errinvt},
the errors from the simulations
agreed well those from the theory.

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsigscan.pdf}
  }
  \caption{
    \label{fig:errsigscan}
    Normalized error, $(T + t_0) \, E$,
    versus the width, $\sigma$,
    for the Gaussian
    and the bandpass updating schemes,
    with $t_0$ given by Eq. \eqref{eq:t0_sinc}.
    %
    In the latter case,
    $\sigma$ is given by Eq. \eqref{eq:sigma_equiv}.
    %
    The points were obtained from averaging results
    from 1000 independent simulations;
    the curves are theoretical predictions.
    %
    \note{This figure was produced by
      \texttt{doc/fig/errsigscan.gp}.
      The data were saved under
      \texttt{data/scan}.
    }%
  }
\end{center}
\end{figure}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/gamcmp.pdf}
  }
  \caption{
    \label{fig:gamcmp}
    Integrals of the autocorrelation functions
    from an MD simulation versus those
    from the one-step process.
    %
    The MD simulation was ran for $10^{10}$ steps with
    the step size $\Delta t_\mathrm{MD} = 0.005$.
    %
    The bin size $\Delta z = 2\, \pi /n$ with $n = 100$,
    and the temperature was $1.0$.
    %
    The velocity rescaling thermostat\cite{bussi2007} was
    applied every MD step with viscosity $20$.
    %
    The reflective boundary condition was used.
    %
    \note{The figure was produced by \texttt{doc/fig/gamcmp.gp}.
      The data were saved under \texttt{data/gamcmp}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}

@@

For the single-bin scheme, we have
$\lambda_2 = \cdots = \lambda_n = 1$,
and
$\Gamma_k$'s affect the error
only via the sum, $\sum_{k = 2}^n \Gamma_k$,
as a multiplicative constant.
